Trends from 30-year observations of downward solar irradiance in Thessaloniki, Greece

### 2023-12-19 MDPI_submission.Rmd
Commit 2023-12-19 15:11

## Aggregation of data and statistical approach

@@

The linear trends were calculated using an first order autoregressive model
(arima) with lag-1, and were reported together with the $2\sigma$ errors (Table\ \ref{tab:trendtable}).

<!-- [For the trends derived from monthly mean data we used an effective -->
<!-- number of observations to calculate the standard error (Sander et al., -->
<!-- 2000) in order to assess the significance of the trends.]{.mark} -->
@Santer2000

@@

## some vars to use later
alltrend   <- signif(LT[DATA == "All sky cond.",    slope], 2)
cleartrend <- signif(LT[DATA == "Clear sky cond.",  slope], 1)
cloudtrend <- signif(LT[DATA == "Cloudy sky cond.", slope], 2)
LTminSSig  <- signif(min(LT$slope.stat_sig), 4)

@@

signif(LT1$`Trend [%/year]` , 3)
round(LT1$`Trend [%/year]` , 3)
format(LT1$`Trend [%/year]` , 3, drop0trailing = T)
as.character(format(LT1$`Trend [%/year]` , 3))

@@

We calculated the linear trends of SDR, from the departures of the mean daily values from the daily climatology and for the three sky conditions (Table\ \ref{tab:trendtable}).
In Figure\ \ref{fig:trendALL} we present only the time series under all-sky conditions; the plots for clear-sky and cloud-sky conditions, are shown in the Appendix (Figures\ \ref{fig:trendCLEAR} and\  \ref{fig:trendCLOUD}).
In the studied period, there is no
significant break or change in the variability pattern of the time
series.
The linear trends in all three datasets are positive and around
$0.4\,\%/y$ for all-sky and cloudy-sky conditions, while for clear-skies the
trend is much smaller (\~$`r cleartrend`\,\%/y$). 
The linear trends were calculated
taking into account the autocorrelation of the time series and all three
are statistically significant at least at the $95\,\%$ confidence level, as
they are larger than the corresponding $2\sigma$ errors.
The clear-sky trend is
very small suggesting a small effect from aerosols and water vapor which
are the dominant factors of the SDR variability. 
In contrast the large
positive trend of SDR under cloudy skies can be attributed to reduction
in cloud cover and/or cloud optical depth. 
Lack of continuous cloud
observations that could support these findings does not allow drawing
firm conclusions. However, there are indications that the total
cloud-cover as inferred from the ERA5 analysis for the grid point of
Thessaloniki is decreasing over the period of study.

@@

In order to unveil further the features of the variability of the three
datasets, Figure\ \ref{fig:cusumnotrendmonthly}
of the CUSUMs, we created another set of CUSUM plots by subtracting the corresponding long term trend from the SDR anomaly data, prior to the CUSUM calculation (Figure\ \ref{fig:cusumnotrendmonthly}).
With this approach periods when the CUSUMs diverge from zero can be interpreted as a systematic variation of SDR from the climatological mean. 
When the CUSUM is increasing, the added values are above the climatological values of the SDR trend and vice versa.
Overall, for all- and clear-sky conditions (Figures\ \ref{fig:cusumnotrendmonthly-1} and\ \ref{fig:cusumnotrendmonthly-2}) we observe periods when the anomalies diverge from the climatological value, each lasting for several years.
The pattern in both datasets is very similar, suggesting prevalence in clear skies over Thessaloniki.
It is interesting that in the period 1993 - 2016 the anomalies have a high variability around zero, while after 2016, the range of the variability is decreased to about one third of the prior period.
For cloudy-sky conditions (Figure\ \ref{fig:cusumnotrendmonthly-3}) the period 1997 - 2008 is dominated by positive CUSUMs, suggesting a reduced effect of clouds on SDR.
From 1997 to mid-2000s CUSUMs are increasing, likely due to a continuous decrease in the optical thickness of clouds, followed by a period of rapid increase (within 3 years) in cloud optical thickness lasting up to 2008.
The following stable period spans for about 15 years up to 2021 when CUSUMs start increasing again.

@@

For all-sky conditions the brightening effect of SDR (positive trend) is
stronger for large SZAs (Figures\ \ref{fig:szatrends-1} and \ref{fig:szatrends-2})
ranging from about $0.1\,\%/y$ to
about $0.7\,\%/y$.
The trends in the morning and afternoon hours are more or
less consistent with small differences at small SZAs which can be
attributed to effects on clear sky SDR from systematic diurnal patterns
of aerosols during the warm period of the year [@Wang2021]. 
Note that SZAs less than $35^\circ$ can only occur during the warm period of the
year when clear skies are more frequent.

@@

Similarly to the long term trends from daily means of SDR discussed above, we have calculated the trend for the three sky conditions and for each season of the year, using the corresponding mean monthly anomalies
(Figure\ \ref{fig:seasonalALL} and Table\ \ref{tab:trendseasontable}).
For all-sky conditions the trend in SDR in winter is the largest
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$),
followed by the trend in autumn
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$,
a value close to the long term trend) both statistically significant at the $99\,\%$ confidence level.
In spring and summer, the trends are much smaller and of lesser statistical significance.
These seasonal differences indicate a possible relation of the trends in SDR to trends in clouds during winter and autumn.
For clear-skies, the trend in winter is $`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Clear skies"], 2)`\,\%/y$, larger than for all-skies ($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$), which is another indication of a decreasing trend in cloud optical thickness.
Moreover, the trends under clear- and cloudy-sky conditions are almost complementary to each other, particularly for winter and autumn, where the signal is stronger.
During spring and summer the statistical significance is very low and the actual trend too small for a meaningful comparison.

@@

We have analyzed a 30 year dataset of SRD measurements in Thessaloniki Greece 
(`r paste(year(min(ALL_1_daily_DESEAS$Date)), " -- ", year(max(ALL_1_daily_DESEAS$Date)))`) aiming to identify the long term variability of
solar irradiance under different sky conditions.
Under all-sky conditions 
there is a positive trend in SDR of $`r alltrend`\,\%/y$ (brightening). 
A previous study [@Bais2013] for the period 1993 -- 2011 reported also a positive trend, of $0.33\,\%/y$.
The slight increase of this trend indicates that the brightening of SDR continues and is probably caused by continuing decreases in aerosol optical depth and the optical thickness of clouds over the area.
A smaller trend has
been found under clear-sky conditions ($`r cleartrend`\,\%/y$) that further supports the assumption that the brightening is caused mainly by decreasing aerosols.
Unfortunately, for the entire period there is no available data for the aerosols, in order to quantify their effect on SDR. 
However, @Siomos2020 have shown that aerosol optical depth over Thessaloniki is decreasing constantly at least up to 2018.
The attenuation of SDR by aerosols over Europe have been proposed as major factor by @Wild2021.
The dimming effect on SDR under cloudy-sky conditions ($`r cloudtrend`\,\%/y$), suggests that cloud optical thickness is decreasing during this period.
Because we have no adequate data to investigate the long term changes of cloud thickness in the region, we cannot verify if the negative SDR trend we observe under under cloudy-skies can be attributed solely to changes in clouds.

The observed brightening on SDR over Thessaloniki is dependent on SZA (larger SZAs lead to stronger brightening).
The trend is also dependent on season, with winter showing the strongest statistically significant trend of
$`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`$
and
$`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Clear skies"], 2)`\,\%/y$
for all- and clear-skies, respectively, in contrast to spring and summer.
The trends for autumn are also significant but smaller (
$`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "All skies"], 2)`$
and
$`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "Clear skies"], 2)`\,\%/y$
for all- and clear-skies, respectively).
Our findings are in agreement with other studies for the region.

>>>>

## Aggregation of data and statistical approach {#aggregationstatistical}

@@

The linear trends were calculated using an first order autoregressive model
 with lag of 1 day with fitting method of the "maximum likelihood" [@Gardner1980; @Jones1980], by implementing the function "arima" from the library "stats" of the R programming language [@RCT2023]. The trends were reported together with the $2\sigma$ errors (Table\ \ref{tab:trendtable}).

<!-- [For the trends derived from monthly mean data we used an effective -->
<!-- number of observations to calculate the standard error (Sander et al., -->
<!-- 2000) in order to assess the significance of the trends.]{.mark} -->
<!-- @Santer2000 -->

@@

## some vars to use later
alltrend   <- signif(LT[DATA == "All sky cond.",    slope], 2)
cleartrend <- signif(LT[DATA == "Clear sky cond.",  slope], 2)
cloudtrend <- signif(LT[DATA == "Cloudy sky cond.", slope], 2)
LTminSSig  <- signif(min(LT$slope.stat_sig), 4)

@@

# signif(LT1$`Trend [%/year]` , 3)
# round(LT1$`Trend [%/year]` , 3)
# format(LT1$`Trend [%/year]` , 3, drop0trailing = T)
# as.character(format(LT1$`Trend [%/year]` , 3))

@@

We calculated the linear trends of SDR, from the departures of the mean daily values from the daily climatology and for the three sky conditions (Table\ \ref{tab:trendtable}).
In Figure\ \ref{fig:trendALL} we present only the time series under all-sky conditions; the plots for clear-sky and cloud-sky conditions, are shown in the Appendix (Figures\ \ref{fig:trendCLEAR} and\  \ref{fig:trendCLOUD}).
In the studied period, there is no
significant break or change in the variability pattern of the time
series.
The linear trends in all three datasets are positive and around
$0.4\,\%/y$ for all-sky and cloudy-sky conditions, while for clear-skies the
trend is much smaller (\~$`r signif(cleartrend,1)`\,\%/y$). 
The linear trends were calculated
taking into account the autocorrelation of the time series and all three
are statistically significant at least at the $95\,\%$ confidence level, as
they are larger than the corresponding $2\sigma$ errors.
The clear-sky trend is
very small suggesting a small effect from aerosols and water vapor which
are the dominant factors of the SDR variability. 
In contrast the large
positive trend of SDR under cloudy skies can be attributed to reduction
in cloud cover and/or cloud optical depth. 
Lack of continuous cloud
observations that could support these findings does not allow drawing
firm conclusions. However, there are indications that the total
cloud-cover as inferred from the ERA5 analysis for the grid point of
Thessaloniki is decreasing over the period of study.

@@

In order to unveil further the features of the variability of the three
datasets, Figure\ \ref{fig:cusumnotrendmonthly}
presents another set of CUSUM plots using anomalies
after the long-term linear trend is removed. With this approach, periods
when the CUSUMs diverge from zero can be interpreted as a systematic
variation of SDR from the climatological mean. When the CUSUM is
increasing, the anomalies values are above the climatology and vice
versa. 
Overall, for all- and cloudy-sky conditions (Figures\ \ref{fig:cusumnotrendmonthly-1} and\ \ref{fig:cusumnotrendmonthly-3}) we
observe periods with anomalies diverging from the climatological values,
each lasting for several years. These fluctuations are probably within
the natural variability and no distinct changes are identified. The
pattern in both datasets is similar, suggesting prevalence in cloudy
skies over Thessaloniki.
For clear skies (Figure\ \ref{fig:cusumnotrendmonthly-2}) the distinct change
in 2004 is now clearer. The most likely reason for this change is the
monotonic reduction of aerosols in Thessaloniki. At that year there a
change in the rate of decrease in aerosol optical depth as illustrated
in Figure 7 of @Siomos2020. This abrupt change in CUSUMs lasts
until about 2010 when the anomalies become again variable.

@@

For all-sky conditions the brightening effect of SDR (positive trend) increases with SZAs (Figures\ \ref{fig:szatrends-1})
ranging from about $0.1\,\%/y$ to
about $0.7\,\%/y$ for the statistically significant trends.
The trends in the morning and afternoon hours are more or
less consistent with small differences at small SZAs which can be
attributed to effects on clear sky SDR from systematic diurnal patterns
of aerosols during the warm period of the year, consistently with the results reported for China by @Wang2021. 
Note that SZAs less than $25^\circ$ can only occur during the warm period of the year around noon when clear skies are more
frequent. The increasing of the trend with SZA is likely caused by the
increased attenuation of SDR with SZA. The effect is larger when aerosol
and/or cloud layers are optically thicker, therefore, decreases in
aerosol and clouds through the study period will result in larger
positive trends of SDR at larger SZAs.

Under clear skies (Figures\ \ref{fig:szatrends-2}), the trends are smaller and less variable,
ranging between $0.1$ and $0.15\,\%/y$ up to $77^\circ$ SZA.
At higher SZAs and in
the afternoon hours there a sharp increase in the trend up to $0.3\,\%/y$,
which may have been caused by the long path length of radiation through
the atmosphere as discussed above for the all-sky conditions. The small

@@

Similarly to the long term trends from daily means of SDR discussed above, we have calculated the trend for the three sky conditions and for each season of the year, using the corresponding mean monthly anomalies
(Figure\ \ref{fig:seasonalALL} and Table\ \ref{tab:trendseasontable}).
<!-- all sky -->
For all-sky conditions the trend in SDR in winter is the largest
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$),
followed by the trend in autumn
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$,
a value close to the long term trend) both statistically significant at the $95\,\%$ confidence level.
In spring and summer, the trends are much smaller and of lesser statistical significance.
These seasonal differences indicate a possible relation of the trends in SDR to trends in clouds during winter and autumn.
<!-- clear sky -->
For clear-skies, the trend in winter is $`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Clear skies"], 2)`\,\%/y$, and is 
associated with the decreasing trend in
aerosol optical depth [@Siomos2020], and almost half of that for
all-skies, which is another indication of a decreasing trend in cloud
optical thickness. 
<!-- clear sky -->
In other seasons the clear-sky trend is still very small
(below $0.1\,\%/y$). 
<!-- cloud dsky -->
Finally, for cloud-skies the
winter trend is the largest 
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Cloudy skies"], 2)`\,\%/y$)
and greater than for all-skies,
followed by a much smaller trend in autumn 
($`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "Cloudy skies"], 2)`\,\%/y$).
<!-- ($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`\,\%/y$) -->

The trends under clear- and cloudy-sky conditions are of the same
direction and it would be expected that their sum is similar to the
all-sky trend. This does not happen, especially for winter, likely due
to the way the monthly means for clear and cloudy skies were calculated.
Daily means were calculated only when at least $60\,\%$ of the clear- or
cloudy-sky data were available (see sect. \ref{aggregationstatistical}).

@@

We have analyzed a 30 year dataset of SRD measurements in Thessaloniki Greece 
(`r paste(year(min(ALL_1_daily_DESEAS$Date)), " -- ", year(max(ALL_1_daily_DESEAS$Date)))`) aiming to identify the long term variability of
solar irradiance under different sky conditions.
Under all-sky conditions 
there is a positive trend in SDR of $`r alltrend`\,\%/y$ (brightening). 
A previous study [@Bais2013] for the period 1993 -- 2011 reported also a positive trend, of $0.33\,\%/y$.
The slight increase of this trend indicates that the brightening of SDR continues and is probably caused by continuing decreases in aerosol optical depth and the optical thickness of clouds over the area.
A smaller trend has
been found under clear-sky conditions ($`r cleartrend`\,\%/y$) which supports the
notion that part of the brightening is caused by decreasing aerosols.
@Siomos2020 have shown that aerosol optical depth over
Thessaloniki is decreasing constantly at least up to 2018.
Unfortunately, for this study aerosol data for the entire period were
not available in order to quantify their effect on SDR. The attenuation
of SDR by aerosols over Europe has been proposed as major factor by @Wild2021.
The brightening effect on SDR under cloudy-sky conditions
($`r cloudtrend`\,\%/y$), suggests that cloud optical thickness is also decreasing
during this period.
As long term data of cloud optical thickness are
also not available for the region, we cannot draw quantitative
conclusions.

The observed brightening on SDR over Thessaloniki is dependent on SZA (larger SZAs lead to stronger brightening).
The trend is also dependent on season, with winter showing the strongest statistically significant trend of
$`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "All skies"], 2)`$
and
$`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Cloudy skies"], 2)`\,\%/y$
for all- and cloudy-skies, respectively, in contrast to spring and summer.
The trends for autumn are also significant but smaller (
$`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "All skies"], 2)`$
and
$`r signif(spp$"Trend [%/year]"[spp$Season == "Autumn" & spp$"Sky condition" == "Cloudy skies"], 2)`\,\%/y$
for all- and cloudy-skies, respectively).
Our findings are in agreement with other studies for the region.
The trend for clear skies is largest in winter
($`r signif(spp$"Trend [%/year]"[spp$Season == "Winter" & spp$"Sky condition" == "Clear skies"], 2)`$) 
and negligible in spring, summer and autumn.

### 2023-12-19 MDPI_submission.tex
Commit 2023-12-19 15:11

\hypertarget{aggregation-of-data-and-statistical-approach}{%
\subsection{Aggregation of data and statistical
approach}\label{aggregation-of-data-and-statistical-approach}}

@@

The linear trends were calculated using an first order autoregressive
model (arima) with lag-1, and were reported together with the
\(2\sigma\) errors (Table~\ref{tab:trendtable}).

\citet{Santer2000}

@@

\begin{verbatim}
[1] 0.380 0.097 0.410
\end{verbatim}

\begin{verbatim}
[1] 0.380 0.097 0.410
\end{verbatim}

\begin{verbatim}
[1] "0.38"  "0.097" "0.41" 
\end{verbatim}

\begin{verbatim}
[1] "0.380" "0.097" "0.410"
\end{verbatim}

@@

In order to unveil further the features of the variability of the three
datasets, Figure~\ref{fig:cusumnotrendmonthly} of the CUSUMs, we created
another set of CUSUM plots by subtracting the corresponding long term
trend from the SDR anomaly data, prior to the CUSUM calculation
(Figure~\ref{fig:cusumnotrendmonthly}). With this approach periods when
the CUSUMs diverge from zero can be interpreted as a systematic
variation of SDR from the climatological mean. When the CUSUM is
increasing, the added values are above the climatological values of the
SDR trend and vice versa. Overall, for all- and clear-sky conditions
(Figures~\ref{fig:cusumnotrendmonthly-1}
and~\ref{fig:cusumnotrendmonthly-2}) we observe periods when the
anomalies diverge from the climatological value, each lasting for
several years. The pattern in both datasets is very similar, suggesting
prevalence in clear skies over Thessaloniki. It is interesting that in
the period 1993 - 2016 the anomalies have a high variability around
zero, while after 2016, the range of the variability is decreased to
about one third of the prior period. For cloudy-sky conditions
(Figure~\ref{fig:cusumnotrendmonthly-3}) the period 1997 - 2008 is
dominated by positive CUSUMs, suggesting a reduced effect of clouds on
SDR. From 1997 to mid-2000s CUSUMs are increasing, likely due to a
continuous decrease in the optical thickness of clouds, followed by a
period of rapid increase (within 3 years) in cloud optical thickness
lasting up to 2008. The following stable period spans for about 15 years
up to 2021 when CUSUMs start increasing again.

@@

For all-sky conditions the brightening effect of SDR (positive trend) is
stronger for large SZAs (Figures~\ref{fig:szatrends-1} and
\ref{fig:szatrends-2}) ranging from about \(0.1\,\%/y\) to about
\(0.7\,\%/y\). The trends in the morning and afternoon hours are more or
less consistent with small differences at small SZAs which can be
attributed to effects on clear sky SDR from systematic diurnal patterns
of aerosols during the warm period of the year \citep{Wang2021}. Note
that SZAs less than \(35^\circ\) can only occur during the warm period
of the year when clear skies are more frequent.

@@

Similarly to the long term trends from daily means of SDR discussed
above, we have calculated the trend for the three sky conditions and for
each season of the year, using the corresponding mean monthly anomalies
(Figure~\ref{fig:seasonalALL} and Table~\ref{tab:trendseasontable}). For
all-sky conditions the trend in SDR in winter is the largest
(\(0.7\,\%/y\)), followed by the trend in autumn (\(0.42\,\%/y\), a
value close to the long term trend) both statistically significant at
the \(99\,\%\) confidence level. In spring and summer, the trends are
much smaller and of lesser statistical significance. These seasonal
differences indicate a possible relation of the trends in SDR to trends
in clouds during winter and autumn. For clear-skies, the trend in winter
is \(0.4\,\%/y\), larger than for all-skies (\(0.7\,\%/y\)), which is
another indication of a decreasing trend in cloud optical thickness.
Moreover, the trends under clear- and cloudy-sky conditions are almost
complementary to each other, particularly for winter and autumn, where
the signal is stronger. During spring and summer the statistical
significance is very low and the actual trend too small for a meaningful
comparison.

@@

We have analyzed a 30 year dataset of SRD measurements in Thessaloniki
Greece (1993 -- 2023) aiming to identify the long term variability of
solar irradiance under different sky conditions. Under all-sky
conditions there is a positive trend in SDR of \(0.38\,\%/y\)
(brightening). A previous study \citep{Bais2013} for the period 1993 --
2011 reported also a positive trend, of \(0.33\,\%/y\). The slight
increase of this trend indicates that the brightening of SDR continues
and is probably caused by continuing decreases in aerosol optical depth
and the optical thickness of clouds over the area. A smaller trend has
been found under clear-sky conditions (\(0.1\,\%/y\)) that further
supports the assumption that the brightening is caused mainly by
decreasing aerosols. Unfortunately, for the entire period there is no
available data for the aerosols, in order to quantify their effect on
SDR. However, \citet{Siomos2020} have shown that aerosol optical depth
over Thessaloniki is decreasing constantly at least up to 2018. The
attenuation of SDR by aerosols over Europe have been proposed as major
factor by \citet{Wild2021}. The dimming effect on SDR under cloudy-sky
conditions (\(0.41\,\%/y\)), suggests that cloud optical thickness is
decreasing during this period. Because we have no adequate data to
investigate the long term changes of cloud thickness in the region, we
cannot verify if the negative SDR trend we observe under under
cloudy-skies can be attributed solely to changes in clouds.

The observed brightening on SDR over Thessaloniki is dependent on SZA
(larger SZAs lead to stronger brightening). The trend is also dependent
on season, with winter showing the strongest statistically significant
trend of \(0.7\) and \(0.4\,\%/y\) for all- and clear-skies,
respectively, in contrast to spring and summer. The trends for autumn
are also significant but smaller ( \(0.42\) and \(0.05\,\%/y\) for all-
and clear-skies, respectively). Our findings are in agreement with other
studies for the region.

>>>>

\hypertarget{aggregationstatistical}{%
\subsection{Aggregation of data and statistical
approach}\label{aggregationstatistical}}

@@

The linear trends were calculated using an first order autoregressive
model with lag of 1 day with fitting method of the ``maximum
likelihood'' \citep{Gardner1980, Jones1980}, by implementing the
function ``arima'' from the library ``stats'' of the R programming
language \citep{RCT2023}. The trends were reported together with the
\(2\sigma\) errors (Table~\ref{tab:trendtable}).

@@

@@

In order to unveil further the features of the variability of the three
datasets, Figure~\ref{fig:cusumnotrendmonthly} presents another set of
CUSUM plots using anomalies after the long-term linear trend is removed.
With this approach, periods when the CUSUMs diverge from zero can be
interpreted as a systematic variation of SDR from the climatological
mean. When the CUSUM is increasing, the anomalies values are above the
climatology and vice versa. Overall, for all- and cloudy-sky conditions
(Figures~\ref{fig:cusumnotrendmonthly-1}
and~\ref{fig:cusumnotrendmonthly-3}) we observe periods with anomalies
diverging from the climatological values, each lasting for several
years. These fluctuations are probably within the natural variability
and no distinct changes are identified. The pattern in both datasets is
similar, suggesting prevalence in cloudy skies over Thessaloniki. For
clear skies (Figure~\ref{fig:cusumnotrendmonthly-2}) the distinct change
in 2004 is now clearer. The most likely reason for this change is the
monotonic reduction of aerosols in Thessaloniki. At that year there a
change in the rate of decrease in aerosol optical depth as illustrated
in Figure 7 of \citet{Siomos2020}. This abrupt change in CUSUMs lasts
until about 2010 when the anomalies become again variable.

@@

For all-sky conditions the brightening effect of SDR (positive trend)
increases with SZAs (Figures~\ref{fig:szatrends-1}) ranging from about
\(0.1\,\%/y\) to about \(0.7\,\%/y\) for the statistically significant
trends. The trends in the morning and afternoon hours are more or less
consistent with small differences at small SZAs which can be attributed
to effects on clear sky SDR from systematic diurnal patterns of aerosols
during the warm period of the year, consistently with the results
reported for China by \citet{Wang2021}. Note that SZAs less than
\(25^\circ\) can only occur during the warm period of the year around
noon when clear skies are more frequent. The increasing of the trend
with SZA is likely caused by the increased attenuation of SDR with SZA.
The effect is larger when aerosol and/or cloud layers are optically
thicker, therefore, decreases in aerosol and clouds through the study
period will result in larger positive trends of SDR at larger SZAs.

Under clear skies (Figures~\ref{fig:szatrends-2}), the trends are
smaller and less variable, ranging between \(0.1\) and \(0.15\,\%/y\) up
to \(77^\circ\) SZA. At higher SZAs and in the afternoon hours there a
sharp increase in the trend up to \(0.3\,\%/y\), which may have been
caused by the long path length of radiation through the atmosphere as
discussed above for the all-sky conditions. The small

@@

Similarly to the long term trends from daily means of SDR discussed
above, we have calculated the trend for the three sky conditions and for
each season of the year, using the corresponding mean monthly anomalies
(Figure~\ref{fig:seasonalALL} and Table~\ref{tab:trendseasontable}). For
all-sky conditions the trend in SDR in winter is the largest
(\(0.7\,\%/y\)), followed by the trend in autumn (\(0.42\,\%/y\), a
value close to the long term trend) both statistically significant at
the \(95\,\%\) confidence level. In spring and summer, the trends are
much smaller and of lesser statistical significance. These seasonal
differences indicate a possible relation of the trends in SDR to trends
in clouds during winter and autumn. For clear-skies, the trend in winter
is \(0.4\,\%/y\), and is associated with the decreasing trend in aerosol
optical depth \citep{Siomos2020}, and almost half of that for all-skies,
which is another indication of a decreasing trend in cloud optical
thickness. In other seasons the clear-sky trend is still very small
(below \(0.1\,\%/y\)). Finally, for cloud-skies the winter trend is the
largest (\(0.76\,\%/y\)) and greater than for all-skies, followed by a
much smaller trend in autumn (\(0.19\,\%/y\)).

The trends under clear- and cloudy-sky conditions are of the same
direction and it would be expected that their sum is similar to the
all-sky trend. This does not happen, especially for winter, likely due
to the way the monthly means for clear and cloudy skies were calculated.
Daily means were calculated only when at least \(60\,\%\) of the clear-
or cloudy-sky data were available (see sect.
\ref{aggregationstatistical}).

@@

We have analyzed a 30 year dataset of SRD measurements in Thessaloniki
Greece (1993 -- 2023) aiming to identify the long term variability of
solar irradiance under different sky conditions. Under all-sky
conditions there is a positive trend in SDR of \(0.38\,\%/y\)
(brightening). A previous study \citep{Bais2013} for the period 1993 --
2011 reported also a positive trend, of \(0.33\,\%/y\). The slight
increase of this trend indicates that the brightening of SDR continues
and is probably caused by continuing decreases in aerosol optical depth
and the optical thickness of clouds over the area. A smaller trend has
been found under clear-sky conditions (\(0.097\,\%/y\)) which supports
the notion that part of the brightening is caused by decreasing
aerosols. \citet{Siomos2020} have shown that aerosol optical depth over
Thessaloniki is decreasing constantly at least up to 2018.
Unfortunately, for this study aerosol data for the entire period were
not available in order to quantify their effect on SDR. The attenuation
of SDR by aerosols over Europe has been proposed as major factor by
\citet{Wild2021}. The brightening effect on SDR under cloudy-sky
conditions (\(0.41\,\%/y\)), suggests that cloud optical thickness is
also decreasing during this period. As long term data of cloud optical
thickness are also not available for the region, we cannot draw
quantitative conclusions.

The observed brightening on SDR over Thessaloniki is dependent on SZA
(larger SZAs lead to stronger brightening). The trend is also dependent
on season, with winter showing the strongest statistically significant
trend of \(0.7\) and \(0.76\,\%/y\) for all- and cloudy-skies,
respectively, in contrast to spring and summer. The trends for autumn
are also significant but smaller ( \(0.42\) and \(0.19\,\%/y\) for all-
and cloudy-skies, respectively). Our findings are in agreement with
other studies for the region. The trend for clear skies is largest in
winter (\(0.4\)) and negligible in spring, summer and autumn.
